Inferring the distribution and rate of Earth-like exoplanets \\ from noisy individual-planet inferences

### 2014-03-27 document/ms.tex
corrected description of uncertainties

\newcommand{\dd}{\ensuremath{\,\mathrm{d}}}
\newcommand{\bvec}[1]{\ensuremath{\boldsymbol{#1}}}

@@

In this \paper, we will assume that the completeness function of the survey is
known and it can be represented as a probability at every (true) period and
radius $\completeness (\entry)$.
Similarly, we can consider the marginalized transit probability $\transitprob
(\period)$ as only a function of period.
These selection effects can be incorporated into the rate from \eq{true-rate}
to get the \emph{observable rate}
\begin{eqnarray}\eqlabel{obs-rate}
\obs{\rate}_\ratepars (\entry)
&=& \rate_\ratepars(\entry)\,\completeness(\entry)\,
\transitprob(\period)\quad.
\end{eqnarray}

Now we will make the further assumption that every candidate in our sample of
detected exoplanets is an independent Poisson sample from the population
described by our rate function.
Under this assumption, the likelihood of a catalog \catalog\ conditioned on a
set of parameters \ratepars\ is (see Appendix XXXXX for a derivation)
\begin{eqnarray}\eqlabel{true-like}
\ln p(\catalog\,|\,\ratepars) &\propto&
\sum_{k=1}^K \ln\obs{\rate}_\ratepars (\entry_k)
- \int \obs{\rate}_\ratepars (\entry) \dd\entry
\quad.
\end{eqnarray}

@@

The ``likelihood function'' given in \eq{true-like}\ldots

In the previous section, we implicitly assumed that the measurements of
\period\ and \radius\ given by the catalog were perfect, without any
observational uncertainties.
This isn't a terrible approximation for the period measurements but not for
the radii.
Luckily, it is not hard to include these effects in our probabilistic
framework.

In general, catalogs produce posterior probability distributions for
quantities conditioned on data.
In particular, the catalog that we're considering here provides estimates of
the posterior distribution of \period\ and \radius\ conditioned on the
observables $\obs{\period}$ and $\obs{\radius}$ under some \emph{interim
priors} parameterized by \interim
\begin{eqnarray}
p(\period,\,\radius\,|\,\obs{\period},\,\obs{\radius},\,\interim) \quad.
\end{eqnarray}
This estimate can be propagated as \emph{posterior samples}
\begin{eqnarray}
\{\period^{(n)},\,\radius^{(n)}\} &\sim&
p(\period,\,\radius\,|\,\obs{\period},\,\obs{\radius},\,\interim) \quad.
\end{eqnarray}

Now the quantity that we need to compute is again the (marginalized)
likelihood of the data $\obs{\period}$ and $\obs{\radius}$ conditioned on a
set of rate function parameters \ratepars
\begin{eqnarray}
p(\obs{\period},\,\obs{\radius}\,|\,\ratepars) &=&
\int p(\period,\,\radius,\,\obs{\period},\,\obs{\radius}\,|\,\ratepars)
\dd\period\dd\radius \\
&=&
\int p(\period,\,\radius\,|\,\ratepars) \,
p(\obs{\period},\,\obs{\radius}\,|\,\period,\,\radius)
\dd\period\dd\radius \\
&=&
\int p(\period,\,\radius\,|\,\ratepars) \,
p(\obs{\period},\,\obs{\radius}\,|\,\period,\,\radius)\,
\frac{p(\period,\,\radius\,|\,\obs{\period},\,\obs{\radius},\,\interim)}
     {p(\period,\,\radius\,|\,\obs{\period},\,\obs{\radius},\,\interim)}
\dd\period\dd\radius \\
&=&
p(\obs{\period},\,\obs{\radius}\,|\,\interim)\,
\int
\frac{p(\period,\,\radius\,|\,\ratepars) \,
      p(\obs{\period},\,\obs{\radius}\,|\,\period,\,\radius)}
     {p(\period,\,\radius\,|\,\interim) \,
      p(\obs{\period},\,\obs{\radius}\,|\,\period,\,\radius)}\,
p(\period,\,\radius\,|\,\obs{\period},\,\obs{\radius},\,\interim)
\dd\period\dd\radius \\
\frac{p(\obs{\period},\,\obs{\radius}\,|\,\ratepars)}
     {p(\obs{\period},\,\obs{\radius}\,|\,\interim)}
&=&
\int
\frac{p(\period,\,\radius\,|\,\ratepars)}
     {p(\period,\,\radius\,|\,\interim)} \,
p(\period,\,\radius\,|\,\obs{\period},\,\obs{\radius},\,\interim)
\dd\period\dd\radius \\
&\approx&
\frac{1}{N}
\sum_{n=1}^N
\frac{p(\period^{(n)},\,\radius^{(n)}\,|\,\ratepars)}
     {p(\period^{(n)},\,\radius^{(n)}\,|\,\interim)}
\end{eqnarray}

>>>>

\newcommand{\dd}{\ensuremath{\,\mathrm{d}}}
\newcommand{\bvec}[1]{\ensuremath{\boldsymbol{#1}}}
\newcommand{\appropto}{\mathrel{\vcenter{
  \offinterlineskip\halign{\hfil$##$\cr
    \propto\cr\noalign{\kern2pt}\sim\cr\noalign{\kern-2pt}}}}}

@@

In this \paper, we will assume that the completeness function of the survey is
known and it can be represented as a probability at every (true) period and
radius $\completeness (\entry)$.
This probability will also include the geometric transit probability as a
function of period $\transitprob (\period)$.
These selection effects can be incorporated into the rate from \eq{true-rate}
to get the \emph{observable rate}
\begin{eqnarray}\eqlabel{obs-rate}
\obs{\rate}_\ratepars (\entry)
&=& \rate_\ratepars(\entry)\,\completeness(\entry)\quad.
\end{eqnarray}

Given this parameterized rate $\obs{\rate}_\ratepars$, we can compute the
Poisson likelihood of a \catalog\ (see Appendix XXXXX for the derivation)
\begin{eqnarray}\eqlabel{true-like}
\ln p(\catalog\,|\,\ratepars) &\propto&
\sum_{k=1}^K \ln\obs{\rate}_\ratepars (\entry_k)
- \int \obs{\rate}_\ratepars (\entry) \dd\entry
\end{eqnarray}
where we have assumed that every object in \catalog\ is an independent
Poisson sample from the rate.

If we had perfect noiseless measurements of the periods and radii for every
object in our catalog, \eq{true-like} provides a metric for determining
constraints on the parameters of the rate function.
In practice, the measurements of \catalog\ are noisy so we need to include
these effects.

@@

The ``likelihood function'' given in \eq{true-like} is a model of the
probability of observing a catalog with \emph{true} periods and radii.
If we want to include the effects of measurement uncertainties, we need to
take a moment to consider the meaning of measurement uncertainties.

\dfmtodo{Introduce the idea of catalogs with uncertainties being a description
of a posterior sampling over catalogs: $\catalog^{(n)} \sim
p(\catalog\,|\,\obs{\catalog},\,\interim)$.}

The quantity that we need to compute is the \emph{marginalized likelihood} of
the data \obs{\catalog}\ given a particular setting of the rate function
parameters \ratepars.
This is achieved by marginalizing over the true periods and radii
\begin{eqnarray}
p(\obs{\catalog}\,|\,\ratepars) &=&
\int p(\catalog,\,\obs{\catalog}\,|\,\ratepars) \dd\catalog \\
&=&
\int p(\catalog\,|\,\ratepars)\,p(\obs{\catalog}\,|\,\catalog) \dd\catalog
\end{eqnarray}
where the second line is true because the rate function is defined on the true
parameters.
This integral can then be written in terms of the sampling under interim
priors
\begin{eqnarray}
p(\obs{\catalog}\,|\,\ratepars)
&=&
\int p(\catalog\,|\,\ratepars)\,p(\obs{\catalog}\,|\,\catalog)\,
\frac{p(\catalog\,|\,\obs{\catalog},\,\interim)}
     {p(\catalog\,|\,\obs{\catalog},\,\interim)}
\dd\catalog
\\
&=&
p(\obs{\catalog}\,|\,\interim)\,
\int \frac{p(\catalog\,|\,\ratepars)}{p(\catalog\,|\,\interim)}\,
p(\catalog\,|\,\obs{\catalog},\,\interim)
\dd\catalog \quad.
\end{eqnarray}
We can approximate this integral using the samples $\catalog^{(n)}$ defined
above and find the marginalized likelihood up to an unimportant normalization
constant
\begin{eqnarray}\eqlabel{marg-like}
p(\obs{\catalog}\,|\,\ratepars) &\appropto&
\sum_{n=1}^N
\frac{p(\catalog^{(n)}\,|\,\ratepars)}{p(\catalog^{(n)}\,|\,\interim)}
\end{eqnarray}
where the numerator inside the sum is given by \eq{true-like} and the
denominator is the interim prior evaluated at the samples.
